**Microbiota predict _Clostridium difficile_ severity in germ-free mice colonized with human feces**

### 2016-12-06 submission/Manuscript.Rmd
citations complete

Murine models to study CDI typically rely on treating conventionally-raised mice with antibiotics either in drinking water or by injection to induce susceptibility [@Buffie2011;@Theriot2011]. This model provides a convenient way to study _C. difficile_ pathogenesis and virulence factors. Numerous microbiome studies have been performed using this model to determine the antibiotic classes (5), starting microbial community (6) and metabolites (7) that impact development and severity of _C. difficile_ infection. While informative, these studies are somewhat removed from human disease because they only examine mouse-associated microbial communities.

Gnotobiotic or germ-free mouse models have been used for a range of studies of CDI, including assessment of species-specific interactions between _C. difficile_ and competing microbial community members (8), analysis of nutrient restriction (9), _in vivo_ transcriptomics of _C. difficile_ and examination of host immune response to CDI (10). Further, CDI therapeutics such as antibiotics and fecal microbiota transplants have been tested extensively in a gnotobiotic-piglet or piglet-to-gnotobiotic-mouse model of disease (11), (12). Pigs have a longer digestive tract with components more similar to humans than mice and are typically infected by strains typical in human infection (13). However, the murine and porcine microbiomes typically do not resemble those of the human gut.

The power of the gnotobiotic models to study CDI has been further realized by first inoculating germ-free mice and piglets with human stool microbes. In one study, germ-free piglets were acutely colonized with human feces for one week and then treated with tigecycline. After challenge with _C. difficile_ none of the antibiotic-treated piglets succumbed to infection, while some of the untreated human-colonized pigs did (11). In another study, germ-free mice colonized with human feces were bred over several generations to create a cohort of mice with identical human-derived microbiomes (14). These mice were subsequently treated with a five-antibiotic cocktail to induce dysbiosis and then were successfully colonized by _C. difficile_ (14).  While informative, these studies were limited in their use of only one human donor as input inoculum.

@@

**Nick will be updating the next two sections**
**Microbes present in the gut prior to infection are predictive of _C. difficile_ colonization levels** Previous work in our group has demonstrated that the microbes present in the mouse microbiome prior to _C. difficile_ challenge can predict future colonization levels (cite Alyx). To determine if the human-associated microbiome in the mouse could similarly predict _C. difficile_ colonization we employed a Random Forest machine learning algorithm similar to the one used in previous studies. Relative abundance of OTUs with greater than 1% abundance at day 0 were used as input to the model. We found that the model explained X% of the variance in colonization levels of _C. difficile_ (Fig 3A). By refining the model to include only the top X predictive OTUs, the model explained X% of the variance. Partial dependency plots of the top X OTUs predictive of _C. difficile_ colonization reveal that colonization is positively associated with _Bacteriodetes_, _Parabacteriodes_ and _Porphyromonadaceae_ (really?) (Fig 3B). Colonization was negatively associated with the presence of _Lachnospiraceae_ (Fig 3B) though none of the human-associated mouse communities were completely resistant to colonization by _C. difficile_ (Fig 2A, 3B). These results confirm that as in mouse microbiome studies, the human-associated gut community prior to challenge is predictive of _C. difficile_ colonization levels.

**The microbiome predicts _C. difficile_ infection severity.** As _C. difficile_ colonization levels can be predicted by the microbiome composition on day 0, we postulated that severity of disease could also be predicted from day 0. Using weight loss as a proxy for severity, we built a Random Forest classification model based on relative abundance to determine which community members were associated with either mild or severe disease. We optimized the model using the AUC-RF algorithm to generate a Reciever Operating Curve (ROC) with maximal Area Under the Curve (AUC) (cite Calle 2011), as used previously in our group to classify colon cancer outcomes (cite Niel 2015 gen med). The optimized model used X number of OTUs and identified several that predicted increased weightloss and subsequent disease severity (Fig 4A, 4B). The populations that were most commonly found in mice that succumbed to _C. difficile_ infection were members of the _Clostridiales_, _Bacilliales_, _Streptococcus_ (Lactobacilles?) and _Bacteriodes_. (need more OTUs here, or number that were used for model and then group by rank). The AUC was calculated using the Out-Of-Bag (OOB) error for each sample (Fig 4A). Cross-validation using 10-fold, leave-one-mouse-out and leave-one-cage-out resulted in AUCs not significantly different from the OOB (Fig 4A). This model reveals that the levels of X OTUs on day 0 can predict disease outcome in mice with human-associated microbiomes. (probably also going to want a supplemental table for predictive OTUs for this and Fig 3)

**Susceptibility to infection is community-dependent.** _C. difficile_ strains are known to vary in disease phenotypes based both on pathogenicity factors of the strain and host microbiome composition (cite someone). To test the susceptibility of human-derived microbiomes to _C. difficile_ infection, two donor-mouse microbiomes that were previously shown to be resistant to _C. difficile_ strain 430 induced weight loss (Fig 2B) (DA00396, DA00430-wait isnt this a control) and one that was susceptible to severe disease (DA00578) were independently challenged with one of three unique _C. difficile_ strains (*table or reference for toxin prod*, etc). As done previously, _C. difficile_ stool CFU was enumerated daily and mouse weight-loss tracked. Infection of resistant mice with virulent strain 458 or the weaker strain 299 resulted in colonization to levels near that of 431 (Fig 5A). Accordingly, these mice did not experience severe weight loss or show clinical signs of disease (Fig 5B). In contrast, infection of susceptible mice (DA00578, Fig 1B) with pathogenic strain 395 resulted in colonization and severe weight loss by two days post-infection (Fig 5A, 5B). Strain 395 is similarly virulent as strain 431, and shows a similar infection phenotype in these mice. These preliminary results suggest that human-associated microbiomes may be more predictive of disease severity than the _C. difficile_ strain.

@@

Study of human-associated microbiomes and their interactions with _C. difficile_ has been limited by the lack of an appropriate model and available samples. Here we colonized 16 groups of mice with human donor stool from patients with different clinical profiles or healthy volunteers. This created a diverse set of human-associated mouse microbiomes that allowed for testing _C. difficile_ infection dynamics and community interactions. Our results showed that both healthy and diarrheal-associated human communities were susceptible to severe disease. Additionally, our results confirmed previous findings that resistance to _C. difficile_ disease is dependent on the entire microbial community and not just one or two taxa (cite alyx). Further, we were able to learn classification models that identified specific OTUs within susceptible communities that predispose mice to infection. By subsequently infecting susceptible or resistant communities with varying strains of _C. difficile_ we confirmed that the initial state of the human-associated microbiome is more predictive of disease severity than the strain used.

Here we colonized germ-free mice with a variety of human stool communities and allowed these communities to colonize the mice for 2 weeks, prior to _C. difficile_ challenge. It is important to note that the mouse gut environment is different from that of the human, and as such exerts a selective pressure on the human stool microbial community that stabilizes over the 2 weeks. Thus these mice cannot be considered "humanized" but rather colonized by human-associated microbes. Interestingly, clinical status was not associated with proclivity to severe _C. difficile_ infection as some healthy donor mice were susceptible. A surprising finding was that none of the human-associated mouse communities were completely resistant to colonization by _C. difficile_. This is in contrast to numerous reports of mouse microbiome studies (cite) and the knowledge that in general healthy humans without gut perturbations are resistant to _C. difficile_ infection (cite). On the other hand, it is known that humans, especially health care employees, can be asymptomatic carriers of _C. difficile_ (Furyura-Kanamari). This may explain the fact that the majority of our groups of mice were colonized by _C. difficile_ but did not display outward signs of illness or infection-associated weight loss (Fig 2).

@@

Finally, our pilot experiments testing the efficiency of infection with different _C. difficile_ strains have allowed for testing community versus pathogen factors of infection. Our results are consistent with previous reports that the composition of the gut microbiome determines if an individual is successfully infected or not (cite Alyx and others). _C. difficile_ strains are known to vary in pathogenicity, primarily via the amount of toxin that is produced during infection (cite). To confirm that disease risk is dependent on the microbiome, we challenged susceptibile mice with an additional pathogenic strain of _C. difficile_ and found that the mice indeed fell ill. In contrast, we exposed an asymptomatic community to a relatively weak strain of _C. difficile_ and found that the community could be colonized by this one as well-- but not cause disease. This suggests that this particular colony is prone to colonization by any _C. difficile_ strain. Further experiments will test if infecting susceptible communities with weak strains allows the mice to survive. It has been recently reported that carriage of a non-toxigenic strain of _C. difficile_ can prevent additional colonization by a pathogenic strain, suggesting a role of niche-filling or immunity this pre-colonization provides (cite cite). One interesting experiment would be to recapitulate this finding in human-associated mouse microbiomes by making a susceptible community resistant to infection. Further work with human-associated communities in mice and mathematical modeling will allow for uncovering of complex human-associated disease processes and dynamics as they relate to human _C. difficile_ infection.

@@

**Human samples** Donor stool samples were chosen from a sample bank collected from patients from October 2010 to November 2012 at the University of Michigan. Sample collection methods and study was approved by the Institutional Review Board at the University of Michigan. Stool samples from University of Michigan Health System patients were submitted to the microbiology lab for _C. difficile_ testing. _C. difficile_ presence was measured using WHATEVER- Cdiff quik check? toxin AB?. Samples from these patients were stored in Cary-Blair medium at -80C. All patients consented to participation in the study. Donor stool for this study was selected from one patient that tested positive for _C. difficile_ infection and five patients that presented with diarrhea and tested negative for _C. difficile_. Healthy donor stool was collected from adult volunteers living in the Ann Arbor, MI area. 100ul of stool slurry was gavaged into each mouse.

**_Clostridium difficile_ and sample processing** For all experiments, 100 spores of _Clostridium difficile_ were used to orally infect mice as described previously (cite alyx). The _C. difficile_ strains used were acquired from spore stocks of (WHERE) and comprised of strains 431, 395, 299 and 468 (cite?). After gavage, the remaining inoculum was diluted and plated to confirm infection dose. Bacteria were cultured anaerobically on TCCFA plates and incubated at 37C (cite plate recipe). 
During infection, two fresh stool pellets were collected from each mouse daily. One pellet was resuspended in anaerobic phosphate-buffered saline and diluted and plated on TCCFA plates to enumerate _C. difficile_ CFU. The other pellet was frozen at -80C until the end of the experiment.

**DNA isolation and sequencing** Bacterial DNA was isolated from banked donor stool and each mouse stool pellet using the MO-BIO PowerSoil DNA isolation kit. This DNA was used as template for amplification of the V4 region of the 16S rRNA gene and fragments were sequenced on an Illumina MiSeq as previously described (cite Jim paper).

>>>>

Murine models to study CDI typically rely on treating conventionally-raised mice with antibiotics either in drinking water or by injection to induce susceptibility [@Buffie2011;@Theriot2011]. This model provides a convenient way to study _C. difficile_ pathogenesis and virulence factors. Numerous microbiome studies have been performed using this model to determine the antibiotic classes [@Schubert2015], starting microbial community [@Reeves2011] and metabolites [@Theriot2013] that impact development and severity of _C. difficile_ infection. While informative, these studies are somewhat removed from human disease because they only examine mouse-associated microbial communities.

Gnotobiotic or germ-free mouse models have been used for a range of studies of CDI, including assessment of species-specific interactions between _C. difficile_ and competing microbial community members [@Reeves2012], analysis of nutrient restriction [@Ng2013], _in vivo_ transcriptomics of _C. difficile_ and examination of host immune response to CDI [@Pawlowski2012]. Further, CDI therapeutics such as antibiotics and fecal microbiota transplants have been tested extensively in a gnotobiotic-piglet or piglet-to-gnotobiotic-mouse model of disease [@Kim2014;@Diao2016]. Pigs have a longer digestive tract with components more similar to humans than mice and are typically infected by strains typical in human infection [@Steele2010]. However, the murine and porcine microbiomes typically do not resemble those of the human gut.

The power of the gnotobiotic models to study CDI has been further realized by first inoculating germ-free mice and piglets with human stool microbes. In one study, germ-free piglets were acutely colonized with human feces for one week and then treated with tigecycline. After challenge with _C. difficile_ none of the antibiotic-treated piglets succumbed to infection, while some of the untreated human-colonized pigs did [@Kim2014]. In another study, germ-free mice colonized with human feces were bred over several generations to create a cohort of mice with identical human-derived microbiomes [@Collins2015]. These mice were subsequently treated with a five-antibiotic cocktail to induce dysbiosis and then were successfully colonized by _C. difficile_ [@Collins2015].  While informative, these studies were limited in their use of only one human donor as input inoculum.

@@

**Nick will be updating the next two sections**
**Microbes present in the gut prior to infection are predictive of _C. difficile_ colonization levels** Previous work in our group has demonstrated that the microbes present in the mouse microbiome prior to _C. difficile_ challenge can predict future colonization levels [@Schubert2015]. To determine if the human-associated microbiome in the mouse could similarly predict _C. difficile_ colonization we employed a Random Forest machine learning algorithm similar to the one used in previous studies. Relative abundance of OTUs with greater than 1% abundance at day 0 were used as input to the model. We found that the model explained X% of the variance in colonization levels of _C. difficile_ (Fig 3A). By refining the model to include only the top X predictive OTUs, the model explained X% of the variance. Partial dependency plots of the top X OTUs predictive of _C. difficile_ colonization reveal that colonization is positively associated with _Bacteriodetes_, _Parabacteriodes_ and _Porphyromonadaceae_ (really?) (Fig 3B). Colonization was negatively associated with the presence of _Lachnospiraceae_ (Fig 3B) though none of the human-associated mouse communities were completely resistant to colonization by _C. difficile_ (Fig 2A, 3B). These results confirm that as in mouse microbiome studies, the human-associated gut community prior to challenge is predictive of _C. difficile_ colonization levels.

**The microbiome predicts _C. difficile_ infection severity.** As _C. difficile_ colonization levels can be predicted by the microbiome composition on day 0, we postulated that severity of disease could also be predicted from day 0. Using weight loss as a proxy for severity, we built a Random Forest classification model based on relative abundance to determine which community members were associated with either mild or severe disease. We optimized the model using the AUC-RF algorithm to generate a Reciever Operating Curve (ROC) with maximal Area Under the Curve (AUC) [@Calle2011], as used previously in our group to classify colon cancer outcomes [@Baxter2016]. The optimized model used X number of OTUs and identified several that predicted increased weightloss and subsequent disease severity (Fig 4A, 4B). The populations that were most commonly found in mice that succumbed to _C. difficile_ infection were members of the _Clostridiales_, _Bacilliales_, _Streptococcus_ (Lactobacilles?) and _Bacteriodes_. (need more OTUs here, or number that were used for model and then group by rank). The AUC was calculated using the Out-Of-Bag (OOB) error for each sample (Fig 4A). Cross-validation using 10-fold, leave-one-mouse-out and leave-one-cage-out resulted in AUCs not significantly different from the OOB (Fig 4A). This model reveals that the levels of X OTUs on day 0 can predict disease outcome in mice with human-associated microbiomes. (probably also going to want a supplemental table for predictive OTUs for this and Fig 3)

@@

Study of human-associated microbiomes and their interactions with _C. difficile_ has been limited by the lack of an appropriate model and available samples. Here we colonized 16 groups of mice with human donor stool from patients with different clinical profiles or healthy volunteers. This created a diverse set of human-associated mouse microbiomes that allowed for testing _C. difficile_ infection dynamics and community interactions. Our results showed that both healthy and diarrheal-associated human communities were susceptible to severe disease. Additionally, our results confirmed previous findings that resistance to _C. difficile_ disease is dependent on the entire microbial community and not just one or two taxa [@Schubert2015]. Further, we were able to learn classification models that identified specific OTUs within susceptible communities that predispose mice to infection.

Here we colonized germ-free mice with a variety of human stool communities and allowed these communities to colonize the mice for 2 weeks, prior to _C. difficile_ challenge. It is important to note that the mouse gut environment is different from that of the human, and as such exerts a selective pressure on the human stool microbial community that stabilizes over the 2 weeks. Thus these mice cannot be considered "humanized" but rather colonized by human-associated microbes. Interestingly, clinical status was not associated with proclivity to severe _C. difficile_ infection as some healthy donor mice were susceptible. A surprising finding was that none of the human-associated mouse communities were completely resistant to colonization by _C. difficile_. This is in contrast to numerous reports of mouse microbiome studies [@Schubert2015] and the knowledge that in general healthy humans without gut perturbations are resistant to _C. difficile_ infection [Heinlen2010]. On the other hand, it is known that humans, especially health care employees, can be asymptomatic carriers of _C. difficile_ [@FuruyaKanamori2015]. This may explain the fact that the majority of our groups of mice were colonized by _C. difficile_ but did not display outward signs of illness or infection-associated weight loss (Fig 2).

@@

Further experiments will test if infecting susceptible communities with weak strains allows the mice to survive. It has been recently reported that carriage of a non-toxigenic strain of _C. difficile_ can prevent additional colonization or recurrence by a pathogenic strain, suggesting a role of niche-filling or immunity this pre-colonization provides [@Gerding2015]. One interesting experiment would be to recapitulate this finding in human-associated mouse microbiomes by making a susceptible community resistant to infection. Further work with human-associated communities in mice and mathematical modeling will allow for uncovering of complex human-associated disease processes and dynamics as they relate to human _C. difficile_ infection.

@@

**Human samples** Donor stool samples were chosen from a sample bank collected from patients from October 2010 to November 2012 at the University of Michigan. Sample collection methods and study was approved by the Institutional Review Board at the University of Michigan. Stool samples from University of Michigan Health System patients were submitted to the microbiology lab for _C. difficile_ testing. _C. difficile_ presence was measured using C.Diff Quik Chek Complete Assay (Techlab). Samples from these patients were stored in Cary-Blair medium at -80C. All patients consented to participation in the study. Donor stool for this study was selected from one patient that tested positive for _C. difficile_ infection and five patients that presented with diarrhea and tested negative for _C. difficile_. Healthy donor stool was collected from adult volunteers living in the Ann Arbor, MI area. 100ul of stool slurry was gavaged into each mouse.

**_Clostridium difficile_ and sample processing** For all experiments, 100 spores of _Clostridium difficile_ were used to orally infect mice as described previously [@Schubert2015]. _C. difficile_ strain 431 is an 027 ribotype clinical isolate isolated from a patient with severe CDI [@Carlson2013]. After gavage, the remaining inoculum was diluted and plated to confirm infection dose. Bacteria were cultured anaerobically on TCCFA plates and incubated at 37C [@Bliss1997]. 
During infection, two fresh stool pellets were collected from each mouse daily. One pellet was resuspended in anaerobic phosphate-buffered saline and diluted and plated on TCCFA plates to enumerate _C. difficile_ CFU. The other pellet was frozen at -80C until the end of the experiment.

**DNA isolation and sequencing** Bacterial DNA was isolated from banked donor stool and each mouse stool pellet using the MO-BIO PowerSoil DNA isolation kit. This DNA was used as template for amplification of the V4 region of the 16S rRNA gene and fragments were sequenced on an Illumina MiSeq as previously described [@Kozich2013].
